---
title: "Portfolio Computational Musicology"
author: "Marijn Biekart"
date: "February 2021"
output: html_document
---

## Description of the corpus
The corpus that I use for my portfolio will consist of 100 opera songs and 100 musical songs, collected from pre-existing playlists that are available on Spotify. I use Spotify's *Opera 100: Spotify Picks* playlist as a basis for my opera playlist. This playlist consists of 97 tracks, so I have added 3 tracks manually, based on Spotify's suggestions. For my musical playlist, I use a public playlist called *BROADWAY MUSICALS*, made by Hugo Torres. Out of all musical playlists I could find, I found this to be the most inclusive. Furthermore, I chose to focus on Broadway musicals because they were written to be performed in front of a live audience, just like operas. The original playlist consists of 150 songs, so I manually removed 50 tracks. I chose to remove tracks from musicals which had more tracks in the playlist, in order to create a playlist with as much different musicials as possible. \
I chose this corpus because I have always been fascinated by musicals. More recently, I was introduced to operas and I recognized the same compelling drama I appreciate in musicals. Opera songs and musical songs both mainly serve to tell a story, but have very different styles. I wonder if opera and musical music share certain aspects, because they both have such a strong narrative function.

## Natural comparison points
In comparing opera tracks with musical tracks, I expect to find a difference in tempo and danceability. Furthermore, I wonder if opera songs are sadder than musical songs, which might be reflected in the valence. I am curious to find out if the energy and loudness differ between the groups. I expect the liveness, intrumentalness, and speechiness to be similar, because most songs in the corpus are studio recordings and contain vocals. 

## Weaknesses of the corpus
Because adding music from *every* opera and musical would create a very big corpus, tracks from some operas and musicals are essentially missing (also because most operas and musicals have had a lot of productions with different artists/conductors/musicians). This means that my corpus does not cover the whole genre. Furthermore, Spotify's pre-existing playlists generally include only the well-known (classical) operas and musicals, leaving out smaller productions.

## Typical tracks
  * Habanera – Carmen: for me, this is a typical opera song with very high notes that everyone knows.
  * La donne e mobile – Rigoletto: again, this is a very famous song. I think the grandeur of this song is typical for opera music.
  * One Day More - Les Miserables: this song is very dramatic and has multiple singers, which is typical for musical songs.
  * You Can't Stop The Beat - Hairspray: the happiness and danceability of this song is typical for musical songs.

## Atypical tracks
  * Summertime - Porgy and Bess: this is a jazzy song, which is a different genre than most opera songs.
  * Ride of the Valkyries - Die Walkure: this song has no lyrics, which is atypical for an opera song.
  * Totally Fucked - Spring Awakening: this comes close to a rock song, which is a different genre than most musical songs. 
  * Land of Lola - Kinky Boots: this song has a strong disco vibe, with more use of electronic instruments than the average musical song.

## Downloading the corpus
```{r}
# Load necessary packages
library(tidyverse)
library(spotifyr)
```
```{r}
# Load playlists 
opera_playlist <- get_playlist_audio_features("", "1rM8PQnCL3unpZENj2xdFa")
musical_playlist <- get_playlist_audio_features("", "7A2ifENYPFKuxloYPlYWtZ")

# Combine playlist into one corpus
corpus <-
  bind_rows(
    opera_playlist %>% mutate(category = "Opera Tracks"),
    musical_playlist %>% mutate(category = "Musical Tracks")
  )
```

## Data visualisation
```{r}
ggplot(corpus, aes(energy, valence, size = danceability, color = tempo)) +
  geom_jitter(alpha = 0.3) +
  facet_wrap(~category) +
  theme_minimal() +
  labs(title = "Valence vs Energy for Musical and Opera Tracks",
       x = "Energy",
       y = "Valence",
       size = "Danceability",
       color = "Tempo") +
  scale_x_continuous(limits = c(-0.1, 1),
                     breaks = c(0, 0.25, 0.5, 0.75, 1),
                     minor_breaks = NULL) +
  scale_y_continuous(limits = c(0, 1),
                     breaks = c(0, 0.25, 0.5, 0.75, 1),
                     minor_breaks = NULL) +
  geom_rug(size = 0.1) +
  geom_text(aes(x = energy, y = valence, label = label),
            data = tibble(label = c("Memory (Cats)", "La serva padrona: Intermezzo I"),
                          category = c("Musical Tracks", "Opera Tracks"),
                          energy = c(0.00803, 0.04950),
                          valence = c(0.1220, 0.6450),
                          danceability = c(0.261, 0.6440),
                          tempo = c(74.153, 116.443)),
            size = 3,
            hjust = "left",
            vjust = "bottom",
            nudge_x = -0.05,
            nudge_y = 0.04)
```

```{r}
ggplot(corpus, aes(energy, fill = category)) +
  geom_histogram(position = "dodge", binwidth = 0.1) +
  theme_minimal() +
  labs(title = "Energy of Muscial and Opera Tracks",
       x = "Energy",
       y = "Count")
```
```{r}
ggplot(corpus, aes(valence, fill = category)) +
  geom_histogram(position = "dodge", binwidth = 0.12) +
  theme_minimal() +
  labs(title = "Valence of Musical and Opera Tracks",
       x = "Valence",
       y = "Count")
```

```{r}
ggplot(corpus, aes(danceability, fill = category)) +
  geom_histogram(position = "dodge", binwidth = 0.1) +
  theme_minimal() +
  labs(title = "Danceability of Musical and Opera Tracks",
       x = "Danceability",
       y = "Count")
```
```{r}
ggplot(corpus, aes(tempo, fill = category)) +
  geom_histogram(position = "dodge", binwidth = 50) +
  theme_minimal() +
  labs(title = "Tempo of Musical and Opera Tracks",
       x = "Tempo",
       y = "Count")
```

